---
title: "hurricane_sst.Rmd"
output:
  html_document:
    df_print: paged
---


```{r echo=FALSE}
#############################################################################################
# Author: Luke Miller, Mar 12, 2018
# Modified: Jeffrey Perkel, Mar 15, 2018
#############################################################################################

# install.packages('tidyverse')
# install.packages('sf')
# install.packages('raster')
# install.packages('mapview')

library(tidyverse)
library(sf)
```

The following code block is adapted from http://r-spatial.org/r/2017/08/28/nest.html, a post on "Tidy storm trajectories", by Edzer Pebesma. It uses a dataset of historical Atlantic hurricane storm tracks that is built into the 'dplyr' package. First it pulls in the latitude/longitude values of each storm, then connects them into tracks. Finally, we look exclusively at those storms in the 2005 season. 

```{r}

storms
storms.sf <- storms %>% st_as_sf(coords = c("long", "lat"), crs = 4326)

storms.sf <- storms.sf %>% 
  mutate(time = as.POSIXct(paste(paste(year,month,day, sep="-"), 
                                 paste(hour, ":00", sep = "")))) %>% 
  dplyr::select(-month, -day, -hour)

storms.nest <- storms.sf %>% group_by(name, year) %>% nest

to_line <- function(tr) st_cast(st_combine(tr), "LINESTRING") %>% .[[1]]

# crs = coordinate reference system, based on EPSG codes
tracks <- storms.nest %>% pull(data) %>% purrr::map(to_line) %>% st_sfc(crs = 4326)

# combining these storm-based geometries to the storm-based attributes:
storms.tr <- storms.nest %>% dplyr::select(-data) %>% st_sf(geometry = tracks)
storms.tr.2k5 <- subset(storms.tr, year == 2005)

```

This code comes from Luke Miller at San Jose State University in California and is based on his 2014 blog post: https://lukemiller.org/index.php/2014/11/extracting-noaa-sea-surface-temperatures-with-ncdf4/. It retrieves sea surface temperatures for a given latitude/longitude and date. 

```{r echo=FALSE}
#############################################################################################
# Basic OISST data load
library (raster)

# customize for your computer
source ("~/Downloads/NOAA_OISST_ncdf4.R") 

# Extract 2 days of data for the North Atlantic ocean. 
# Customize the file paths to suit your local machine
ssts = extractOISSTdaily("~/Downloads/sst.day.mean.2005.v2.nc","~/Downloads/lsmask.oisst.v2.nc", lonW=160, lonE=359, latS=-10, latN=60, date1='2005-09-01', date2='2005-09-02')

s = brick(ssts, 
          xmn = as.numeric(attr(ssts, 'dimnames')$Long[1]),
          xmx = as.numeric(attr(ssts, 'dimnames')$Long[ncol(ssts)]),
          ymn = as.numeric(attr(ssts, 'dimnames')$Lat[nrow(ssts)]),
          ymx = as.numeric(attr(ssts, 'dimnames')$Lat[1]),
          crs = '+proj=longlat +datum=WGS84')
s = dropLayer(s, 2)

# Necessary steps to get leaflet to plot western hemi data
# See https://github.com/r-spatial/mapview/issues/6
b = shift(s, -360)
SST <- crop(b, extent(-100,180,-90,90),snap='in') %>% trim()

#############################################################################################
```

Plot the data on interactive maps using the mapview and leaflet R packages

```{r}
# Generate the initial map of SSTs, and then add storm tracks 
# Center view on Bermuda
library(leaflet)
library (mapview)

m <- mapview(SST, legend = TRUE, trim = FALSE) + 
  mapview(storms.tr.2k5, zcol = "name", legend = FALSE)
m@map <- m@map %>% setView(-64.768, 32.295, zoom = 3) 
m
```

Plot NHC forecast track for Hurricane Irma (2017) using publicly available data (advisory #20), overlaid with potential coastal targets. 

Irma data source: https://www.nhc.noaa.gov/gis/forecast/archive/al112017_5day_020.zip

```{r echo=FALSE}
library (rgdal)

df <- read.csv(textConnection(
"Name,Lat,Long
Barbuda,17.648,-61.779
Sint Maarten,18.060,-63.049
British Virgin Islands,18.349,-64.718
Santo Domingo,18.483,-69.929
San Juan,18.421,-66.059
Havana,23.111,-82.357
Key West,24.555, -81.779
Miami Beach,25.790,-80.135
New Orleans,29.999,-90.087
Houston,29.782,-95.405
"))

irmaPolygon <- readOGR("/Users/jeffreyperkel/Downloads/al112017_5day_020/al112017-020_5day_pgn.shp")
irmaLine <- readOGR("/Users/jeffreyperkel/Downloads/al112017_5day_020/al112017-020_5day_lin.shp")
irmaPoints <- readOGR("/Users/jeffreyperkel/Downloads/al112017_5day_020/al112017-020_5day_pts.shp")
```

Plot the map using Leaflet. Color the coastal cities in blue, the hurricane in red, and give the polygon a black border. Add text to label the figure. And center it on Havana, Cuba.

```{r}

# convert the date info in the points data into a more useful format, from eg '2017/09/04 02:00:00' to '02:00 AM Tue'
# labels <- format (as.POSIXlt(as.data.frame(irmaPoints)$FLDATELBL, format="%Y/%m/%d %I:%M:%S"), format = "%I:%M %p %a")

labels <- as.data.frame(irmaPoints)$FLDATELBL

m <- leaflet () %>% addTiles()  
m <- m %>% addPolygons(data = irmaPolygon, color = "black", weight = "2", fillColor="red", fillOpacity = 0.3) %>% 
  addPolylines(data = irmaLine) %>% 
  addCircleMarkers(data = irmaPoints, color = "red", radius = 2, label = labels, labelOptions = labelOptions(noHide = TRUE, textOnly = TRUE)) %>% 
  addCircleMarkers(lat=df$Lat,lng=df$Long,label=df$Name, radius = 2, labelOptions = labelOptions(noHide = TRUE, textOnly = TRUE)) %>%
  setView(-82.357,23.111,zoom=4)

m
```

Let's overlay SST data for the same day...

```{r}
ssts = extractOISSTdaily("~/Downloads/sst.day.mean.2017.nc","~/Downloads/lsmask.oisst.v2.nc", lonW=160, lonE=359, latS=-10, latN=60, date1='2017-09-04', date2='2017-09-05')

s = brick(ssts, 
          xmn = as.numeric(attr(ssts, 'dimnames')$Long[1]),
          xmx = as.numeric(attr(ssts, 'dimnames')$Long[ncol(ssts)]),
          ymn = as.numeric(attr(ssts, 'dimnames')$Lat[nrow(ssts)]),
          ymx = as.numeric(attr(ssts, 'dimnames')$Lat[1]),
          crs = '+proj=longlat +datum=WGS84')
s = dropLayer(s, 2)

# Necessary steps to get leaflet to plot western hemi data
# See https://github.com/r-spatial/mapview/issues/6
b = shift(s, -360)
SST <- crop(b, extent(-100,180,-90,90),snap='in') %>% trim()

m <- m %>% addRasterImage (SST, colors = "Set3") 
m <- m %>% setView(-64.768, 32.295, zoom = 4)
m
```

Note that the 'raster' package can also plot the OISST data directly, and provide a nice legend...

```{r}
plot (SST)
```

Alternatively, you can convert shape files to GeoJSON using ogr2ogr web client: https://ogre.adc4gis.com. For this to work, you need to create a ZIP file containing all the elements of the Shape file)


```{r}
library(geojsonio)

# import the GeoJSON files
irmaPolygon <- geojson_read("~/Google_Drive/irma_pgn.json", what = "sp")
irmaLine <- geojson_read("~/Google_Drive/irma_lin.json", what = "sp")
irmaPoints <- geojson_read("~/Google_Drive/irma_pts.json", what = "sp")

#
# NOTE: for some reason, geojson_read() messes up the FLDATELBL import, 
# so I've used the VALIDTIME field instead.
#
labels <- as.data.frame(irmaPoints)$VALIDTIME

m <- leaflet () %>% addTiles()  
m <- m %>% addPolygons(data = irmaPolygon, color = "black", weight = "2", fillColor="red", fillOpacity = 0.3) %>% 
  addPolylines(data = irmaLine) %>% 
  addCircleMarkers(data = irmaPoints, color = "red", radius = 2, label = labels, labelOptions = labelOptions(noHide = TRUE, textOnly = TRUE)) %>% 
  addCircleMarkers(lat=df$Lat,lng=df$Long,label=df$Name, radius = 2, labelOptions = labelOptions(noHide = TRUE, textOnly = TRUE)) %>%
  setView(-82.357,23.111,zoom=4)

m

```

